Data table crashes
========================================================

I am having a similar issue to this post: http://r.789695.n4.nabble.com/fread-crash-td4683394.html

The file is about 550MB, i'm unsure how many rows it actually is (several million).

When i try to run fread, Rstudio just crashes with no error. 
I can read in up to about 15 rows

```{r}

require(data.table)

#env dist table

env<-fread("EnvData.csv",nrows=15,verbose=TRUE)
head(env)
```

However when i run fread with more than 20 rows, it crashes Rstudio. 

```{r eval=FALSE}
#not run
env<-fread("EnvData.csv",nrows=25,verbose=TRUE)
```

*verbose on the error output reads:*

Input contains no \n. Taking this to be a filename to open

File opened, filesize is  0.543B

File is opened and mapped ok

Detected eol as \n only (no \r afterwards), the UNIX and Mac standard.

Using line 30 to detect sep (the last non blank line in the first 'autostart') ... sep=','

Found 4 columns

First row with 4 fields occurs on line 2 (either column names or first row of data)

Some fields on line 2 are not type character (or are empty). Treating as a data row and using default column names.

Count of eol after first data row: 15989212

Subtracted 0 for last eol and any trailing empty lines, leaving 15989212 data rows

nrow limited to nrows passed in (25)

Type codes: 4113 (first 5 rows)

Type codes: 4113 (+middle 5 rows)

*Look at the file, nothing seems wrong*

```{r}

env<-read.csv("EnvData.csv",nrows=25)

env
```

